---
title: "Measuring the Microbiome:</br>Methods & Tools"
subtitle: "📚Epidemiology 674📚"  
author: 
  - "Brendan J. Kelly, MD, MS"
  - "Kyle Bittinger, PhD"
date: 'Updated: `r format(Sys.Date(), "%d %B %Y")`'
output:
  xaringan::moon_reader:
    self_contained: TRUE
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
background-image: url(svg/bacteria.svg)
background-size: 500px
background-position: 50% 50%
class: center, middle, inverse

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE, eval=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#011F5B",
  secondary_color = "#990000",
  inverse_header_color = "#FFFFFF",
  header_font_google = google_font("Roboto Condensed"),
  text_font_google = google_font("Roboto"),
  code_font_google = google_font("Roboto Mono"),
  # padding = "64px 64px 64px 64px"
  # base_font_size = "24px",
  # text_font_base = "1rem",
  # header_h1_font_size = "2.75rem",
  # header_h2_font_size = "2rem",
  # header_h3_font_size = "1.25rem",
)
```

# Welcome!



---

# Where am I? How did I get here?

.pull-left[

<span style="font-size: 1.25rem; text-indent: 100px">Epidemiology 674</span>
    
<span style="font-size: 1.25rem;">Measuring the Microbiome:<br/>Methods & Tools</span>

]

.pull-right[

- You study epidemiology  

- You study microbiology.

- You study informatics.

- You study statistics.

]


---
background-image: url(svg/micro_petri.svg)
background-size: 500px
background-position: 50% 50%
class: center, middle, inverse

# How to measure "the microbiome"?


---

# Membership, genes, or metabolites?

.pull-left[

- Which "microbiome"?

    - microbial ecology

    - microbial metabolism

    - host-microbe interactions

]

.pull-right[

- measurement methods and tools differ __BUT__ data structures similar and create some common concerns

    - sparse count data
    
    - misclassification
    
]








---
background-image: url(svg/dna.svg)
background-size: 500px
background-position: 50% 50%
class: center, middle, inverse

# How are microbiome data structured?



---
background-image: url(svg/stacked_barchart.svg)
background-size: 500px
background-position: 50% 50%
class: center, middle, inverse

# Pitfalls and dangers... 



---
background-image: url(svg/stacked_barchart.svg)
background-size: 500px
background-position: 50% 50%
class: center, middle, inverse

# ... analysis methods and models.  



---
class: inverse, center, middle

# Get Started

---

# Hello World

Install the **xaringan** package from [Github](https://github.com/yihui/xaringan):

```{r eval=FALSE, tidy=FALSE}
devtools::install_github("yihui/xaringan")
```



---
background-image: url(svg/bacteria.svg)
background-size: 100px
background-position: 98% 90%
class: center, middle

# Thank you!
#### Slides available: [github.com/bjklab](https://www.gihub.com/bjklab/microbiome-measures_01_data-and-danger)
#### [brendank@pennmedicine.upenn.edu](brendank@pennmedicine.upenn.edu)




